**Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system**

### 2016-01-05 Schloss_PacBio16S_PeerJ_2016.Rmd
Update code and text for basic error correction section

regions <- c("v4", "v13", "v35", "v15", "v16", "v19")
clrs <- rainbow(length(regions))
pretty.region <- c("v13"="V1-V3", "v15"="V1-V5", "v16"="V1-V6", "v19"="V1-V9", "v35"="V3-V5", "v4"="V4")
names(clrs) <- regions

@@

ins <- rep(as.numeric(rownames(error_quality)), error_quality$insertions)
i.quant <- quantile(ins, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
i.94 <- 100 * sum(ins >= 94)/length(ins)
# [1] 80.35283
```

***The PacBio error profile.***
To build a sequence curation pipeline, we first needed to characterize the error
rate associated with sequencing the 16S rRNA gene. We observed an average
sequencing error rate of `r format(100*init_error_rate, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions
accounted for `r format(insertion_rate, digits=3, nsmall=1)`,
`r format(deletion_rate, digits=3, nsmall=1)`, and `r format(substitution_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally
likely and all four bases
were equally likely to cause insertion errors. Interestingly, guanines
(`r format(deletion_bias["dG"], digits=3, nsmall=1)`%) and adenines
(`r format(deletion_bias["dA"], digits=3, nsmall=1)`%) were more likely to be deleted
than cytosines (`r format(deletion_bias["dC"], digits=3, nsmall=1)`%) or thymidines
(`r format(deletion_bias["dT"], digits=2, nsmall=1)`%). The PacBio quality values varied between
`r paste(range(c(matches, subs, ins)), collapse=' and ')`. Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(m.94, digits=3, nsmall=1)`%), substitutions (`r format(s.94, digits=3, nsmall=1)`%), and insertions (`r format(i.94, digits=3, nsmall=1)`%). Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(cor(mock_error_good$error, mock_error_good$pred_error), digits=2, nsmall=2)`; Figure 1A).

```{r initialError}
getInitError <- function(folder){
	error <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".mock.filter.error.summary", sep=""), header=T, row.names=1)
	error <- error[error$numparents==1,]
	return(sum(error$mismatches)/sum(error$total))
}

init.rates <- unlist(lapply(regions, getInitError))
names(init.rates) <- regions

getReasonsLost <- function(region){
	composite <- read.table(file=paste("pipeline_dev/", region, "/", region, ".composite", sep=""), header=T, row.names=1)
	reasons <- as.character(composite$reason)

	nseqs <- length(reasons)
	good <- sum(reasons=="g")	/ nseqs
	start.stop <- sum(grepl("s", reasons) | grepl("e", reasons)) / nseqs
	homop <- length(grep("h", reasons)) / nseqs
	ambig <- length(grep("n", reasons)) / nseqs
	return(c(good=good, start.stop=start.stop, homop=homop, ambig=ambig, ngood=nseqs))
}

reasons.lost <- matrix(unlist(lapply(regions, getReasonsLost)), ncol=5, byrow=T)
colnames(reasons.lost) <- c("good", "start.stop", "homop", "ambig", "nseqs")
rownames(reasons.lost) <- regions

orig.nseqs <- reasons.lost[,"nseqs"]
basic.nseqs <- orig.nseqs * reasons.lost[,1]
reasons.lost <- reasons.lost[,-5]

getBasicError <- function(folder){
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
	good <- composite[composite$reason == "g",]
	return(sum(good$mismatches)/sum(good$total))
}

basic.error <- unlist(lapply(regions, getBasicError))
names(basic.error) <- regions
#         v4         v13         v35         v15         v16         v19
#0.005373722 0.014743697 0.013730112 0.009027326 0.014249660 0.013027623
composite.basic.error <- 100 * sum(basic.error*basic.nseqs)/(sum(basic.nseqs))
```

***A basic sequence curation procedure.***
To establish a simple curation procedure, we culled any sequence that contained
an ambiguous base call, had a string of the same base repeated 9 or more times,
did not start and end at the expected alignment coordinates for that region
of the 16S rRNA gene, or that was chimeric. This reduced the experiment-wide
error rate from `r format(init_error_rate, digits=3, nsmall=2)` to
`r format(composite.basic.error, digits=2, nsmall=2)`%. This basic procedure
resulted in the removal of between
`r format(round(min(100*(1-reasons.lost[,"good"])), digits=1), nsmall=1)`
(`r pretty.region[names(which.min(100*(1-reasons.lost[,"good"])))]`) and
`r round(max(100*(1-reasons.lost[,"good"])), digits=1)`
(`r pretty.region[names(which.max(100*(1-reasons.lost[,"good"])))]`)% of the
reads. The percentage of reads removed increased with the length of the
fragment (Figure 2). The number of reads removed
because of the presence of ambiguous base calls was similar to the number of
reads that were removed for not fully aligning to the correct region within the
16S rRNA gene (Table 2). The latter class of errors was
generally due to sequence truncations that could not be explained.

>>>>

regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
clrs <- rainbow(length(regions))
pretty_region <- c("V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9", "V3V5"="V3-V5", "V4"="V4")
names(clrs) <- regions

@@

ins <- rep(as.numeric(rownames(error_quality)), error_quality$insertions)
i.quant <- quantile(ins, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
i.94 <- 100 * sum(ins >= 94)/length(ins)
# [1] 80.35283

## Basic error correction analysis
get_raw_error <- function(region){
	region_data <- mock_error_good[mock_error_good$region == region,]
	return(100*sum(region_data$mismatches)/sum(region_data$total))
}

raw_rates <- sapply(regions, get_raw_error)

get_basic_kept <- function(region){
	region_data <- mock_error_good[mock_error_good$region == region,]
	100 * sum(region_data$good_coords & region_data$good_homop)/nrow(region_data)
}

basic_kept <- sapply(regions, get_basic_kept)

get_basic_error <- function(region){
	region_data <- mock_error_good[mock_error_good$region == region & mock_error_good$good_coords & mock_error_good$good_homop,]
	return(100*sum(region_data$mismatches)/sum(region_data$total))
}

basic_rates <- sapply(regions, get_basic_error)

mock_error_basic <- mock_error_good[mock_error_good$good_coords & mock_error_good$good_homop,]

basic_error_rate <- 100 * sum(mock_error_basic$mismatches) / sum(mock_error_basic$total)		#[1] 0.6448736
```

***The PacBio error profile and a basic sequence curation procedure.***
To build a sequence curation pipeline, we first needed to characterize the error
rate associated with sequencing the 16S rRNA gene. We observed an average
sequencing error rate of `r format(100*init_error_rate, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions
accounted for `r format(insertion_rate, digits=3, nsmall=1)`,
`r format(deletion_rate, digits=3, nsmall=1)`, and `r format(substitution_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally
likely and all four bases
were equally likely to cause insertion errors. Interestingly, guanines
(`r format(deletion_bias["dG"], digits=3, nsmall=1)`%) and adenines
(`r format(deletion_bias["dA"], digits=3, nsmall=1)`%) were more likely to be deleted
than cytosines (`r format(deletion_bias["dC"], digits=3, nsmall=1)`%) or thymidines
(`r format(deletion_bias["dT"], digits=2, nsmall=1)`%). The PacBio quality values varied between
`r paste(range(c(matches, subs, ins)), collapse=' and ')`. Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(m.94, digits=3, nsmall=1)`%), substitutions (`r format(s.94, digits=3, nsmall=1)`%), and insertions (`r format(i.94, digits=3, nsmall=1)`%). Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(cor(mock_error_good$error, mock_error_good$pred_error), digits=2, nsmall=2)`; Figure 1A).

We established a simple curation procedure by culling any sequence that had a
string of the same base repeated more than 8 times  or did not start and end at
the expected alignment coordinates for that region of the 16S rRNA gene. This
reduced the experiment-wide error rate from `r format(init_error_rate, digits=3, nsmall=2)` to
`r format(basic_error_rate, digits=2, nsmall=2)`%. This basic procedure
resulted in the removal of between
`r format(min(100-basic_kept), digits=3, nsmall=1)`
(`r pretty_region[names(which.min(100-basic_kept))]`) and
`r format(max(100-basic_kept), digits=3, nsmall=1)`
(`r pretty_region[names(which.max(100-basic_kept))]`)% of the reads (Table 2). Although
the percentage of reads removed increased with the length of the fragment, there
was no obvious relationship between fragment length and error rate (Figure 2).

```{r initialError}

```
